---
title: "Intro to R - Module 3"
output: html_document
date: "2025-05-21"
---

# **Module 3: Plotting with ggplot**

## In this script, we will explore how to use the `ggplot2` to create and customize plots. 

## `gg` stands for "Grammar of Graphics," which is a framework towards producing quantitative graphics espoused in a book by Leland Wilkinson:

## Wilkinson, L. (2005). The Grammar of Graphics. (2 Ed.). Springer. https://www.doi.org/10.1007/0-387-28695-0

## `ggplot` is a package of plotting functions that aims to apply this framework is R:

## https://ggplot2.tidyverse.org/



## Installation

```{r}
#install.packages("ggplot2")
library(ggplot2)
```


## Practice Dataset

```{r}
# We are going to use titanic library to creat plots

# Titanic data
#install.packages("titanic")
library(titanic)

# Tibble
library(dplyr)
titanic <- as_tibble(titanic_train)


# ggplot works in layers
# We start with a blank canvas
ggplot(titanic, aes(x = Age, y = Fare))

# We specified which dataset we want (i.e., titanic) and
# set our X and Y axes.

# Next, we can add some layers using '+' and various other functions
# For example, we can add a layer to add points in the form of (Age,Fare) to this plot.

ggplot(titanic, aes(x = Age, y = Fare))+
  geom_point()

# The layer we just added draws a scatter plot on the canvas that we made earlier.
# We will see additional functions/layers to change to plot title, labels, colors, add more graphics onto the plot, and much more. By going layer by layer, ggplot gives users a lot of flexibility. 


# geom_ ... is a common structure for ggplot's different plot types.
# We find:
# geom_histogram(), geom_boxplot(), geom_bar(), geom_point(), geom_line()
# among other kinds of plot in the package (see cheatsheet for more examples).

# Let's see some examples:
ggplot(titanic, aes(x = Age, y = Fare))+
  geom_line()

ggplot(titanic, aes(x = Age, y = Fare))+
  geom_smooth()

# Some single variable plots:
ggplot(titanic, aes(x = Age))+
  geom_histogram()

ggplot(titanic, aes(x = Age))+
  geom_boxplot()

# Or can flip the boxplot, if you prefer
ggplot(titanic, aes(x = Age))+
  geom_boxplot()+
  coord_flip()

# Using + we can add even more layers and combine plots that we've seen.
ggplot(titanic, aes(x = Age, y = Fare))+
  geom_point() +
  geom_line()

ggplot(titanic, aes(x = Age, y = Fare))+
  geom_point()+
  geom_smooth()

# We can add color to data points based on the levels in another variable:
ggplot(titanic, aes(x = Age, y = Fare, color = Sex))+
  geom_point()+
  geom_smooth()
```


## Data Points

```{r}
# You can change the size of data points:
ggplot(titanic, aes(x = Age, y = Fare, color = as.factor(Survived)))+
  geom_point(size = 2.5)+
  geom_smooth(alpha = .5)

# Or you can change the type of points: 
ggplot(titanic, aes(x = Age, y = Fare, color = Sex))+
  geom_point(shape = 6)+
  geom_smooth()

# See a full list of shape at: 
# https://www.r-bloggers.com/2021/06/r-plot-pch-symbols-different-point-shapes-in-r/
```


## Labels

```{r}
# You can add label to your plots by labs() function as follows:

ggplot(titanic, aes(x = Age, y = Fare, color = Sex))+
  geom_point(shape = 6) +
  labs(title = "Scatterplot of Age and Fare on the Titanic by Sex",
       x = "Age in Years",
       y = "Fare Paid in $",
       subtitle = "Includes all recorded passengers of the Titanic",
       caption = "Created at Data Vis workshop")
```


## Themes

```{r}
# We can control the theme of the graphs
# The layer theme_* can be used for this purpose.


# The classic dark-on-light ggplot2 theme
ggplot(titanic, aes(x = Age, y = Fare, color = Sex))+
  geom_point(shape = 6) +
  labs(title = "Scatterplot of Age and Fare on the Titanic by Sex",
       x = "Age in Years",
       y = "Fare Paid in $",
       subtitle = "Includes all recorded passengers of the Titanic",
       caption = "Created at Data Vis workshop")+
  theme_bw()


# A void theme
ggplot(titanic, aes(x = Age, y = Fare, color = Sex))+
  geom_point(shape = 6) +
  labs(title = "Scatterplot of Age and Fare on the Titanic by Sex",
       x = "Age in Years",
       y = "Fare Paid in $",
       subtitle = "Includes all recorded passengers of the Titanic",
       caption = "Created at Data Vis workshop")+
  theme_light()


# Many more, you can see them at the help documentation. 
# Run help("theme_bw")
```


## Exercise 1

```{r}
# In this exercise, we work with iris dataset available at datasets packages. 
# Take your time and read the help documentation for this dataset. 
# You can find it by running help(iris) or ?iris in the console.
# After that, start working on the following parts.

# Part 1) use ggplot to draw a scatterplot where the length of Sepal is on the x-axis and width of Sepal is on the y-axis.


# Part 2) Color the data points in the plot based on the values in Species column.


# Part 3) Increase the point size to a larger value (your choice of value) and add some label to your plot.


# Part 4) Note that Species column have three unique values 
# (You can verify this by running unique(iris$Species) command line in console)
# Suppose now we want to plot data only for "setosa" and "versicolor".
# Use what we learned in this workshop to: 
# (i) Select the rows that match this condition, using dplyr functionality.
# (ii) Save this in a new object called iris.new
# (iii) Repeat parts 1 to 3.
```


## Changing the colour and label of different categories

```{r}
# No we want to see how we can change the colour and label of different categories of a variable

ggplot(titanic, aes(x = Age, y = Fare, color = Sex))+
  geom_point()+
  geom_smooth()+
  scale_color_manual(values = c("male" = "orange", "female" = "blue"),                      
                     labels = c("female" = "Female", "male" = "Male"))


# Changing the colour in histogram
ggplot(titanic, aes(x = Age)) +
  geom_histogram(col = "black", fill = "darkred")
```


## Rescaling

```{r}
ggplot(titanic, aes(x = Age, y = Fare)) +
  geom_point()+
  scale_x_continuous(limits = c(0,90), breaks = seq(0,90,10)) + 
  scale_y_continuous(limits = c(0,300), breaks = seq(0,300,100))
```


## Export

```{r}
ggplot(titanic, aes(x = Age, y = Fare, color = Sex))+
  geom_point(shape = 1) +
  labs(title = "Scatterplot of Age and Fare on the Titanic by Sex",
       x = "Age in Years",
       y = "Fare Paid in $",
       subtitle = "Includes all recorded passengers of the Titanic",
       caption = "Created at Data Vis workshop")+
  scale_color_manual(values = c("male" = "red", "female" = "blue"),                      
                     labels = c("female" = "Female", "male" = "Male")) -> titanic.plot
  
ggsave("titanicscatterplot.png", titanic.plot, width = 6.5, height = 6.5, units = "in", dpi = 300)
```


## Exercise 2

```{r}
# In this exercise, we work with iris dataset available at datasets packages. 
# Take your time and read the help documentation for this dataset. 
# You can find it by running help(iris) or ?iris in the console.
# After that, start working on the following parts.

# Part 1) use ggplot to draw a scatterplot where the length of Sepal is on the x-axis and width of Sepal is on the y-axis.


# Part 2) Color the data points in the plot based on the values in Species column.


# Part 3) Increase the point size to a larger value (your choice of value) and add some label to your plot.


# Part 4) Note that Species column have three unique values 
# (You can verify this by running unique(iris$Species) command line in console)
# Suppose now we want to plot data only for "setosa" and "versicolor".
# Use what we learned in this workshop to: 
# (i) Select the rows that match this condition, using dplyr functionality.
# (ii) Save this in a new object called iris.new
# (iii) Repeat parts 1 to 3.
```

