---
title: "Intro to R Workshop - Module 1"
output: html_document
date: "2025-05-21"
---

# **Module 1: Data types and Data structures**

## In this first module, we will see different data types that R can handle.

## These data types are stored in different data structures. We will briefly go over a few data structures before getting to the data frame, which is the main structure that we will work with in the workshop.


## **Data Types**

## The most commonly used data types are: (1) numeric or double, (2) strings (i.e., words/characters), and (3) logicals.


## 1. Numeric

```{r}
# We can define variable x to be a number as follows:
x <- 1.2345
x
# We can choose any variable name (with some limits, for example, no spaces):
my_favorite_number <- 2
my_favorite_number

# With numerics, R can perform mathematical operations like +, -, /, *, exp(), log(), etc...
log(2)
x / 2
```

## 2. Strings

```{r}
# These are variables that hold characters. To define a string, we use " ":
day <- "Monday"
day 
weather <- "Rainy :("
weather
```

## 3. Logicals

```{r}
# These work for binary TRUE/FALSE data. They are used to check conditions (e.g., greater than or equal to >=, less than <, etc...)
did_it_rain = TRUE 
did_it_rain

# Not statement
!did_it_rain

did_it_snow = FALSE

# Did it rain and snow? Use the AND operator: &
rain_and_snow <- did_it_rain & did_it_snow
rain_and_snow

# Did it rain OR did it snow? Use OR operator: |
rain_or_snow <- did_it_rain | did_it_snow
rain_or_snow

# We can use == to check for equality:
my_favorite_number == 2

# Note the distinction: "=" assigns the value on the right side to the variable 
# on the left
my_favorite_number == 2
# "==" checks for equality
```


## Exercise 1

```{r}
# (1) Use the less than operator < to verify whether x is less than my_favorite_number
#     (save this result and print it)

# (2) Find the logical variables in the environment and check whether NOT one of them is TRUE

# (3) Define a string variable and print it


```

## There are other data types, such as integers and complex numbers, but we will not be covering them here.

## factors are another data type that is used for categorical data analysis. We will see this later.


## **Data Structures**

## Data Structures are objects used to store elements in R.
## We will introduce quickly vectors, matrices, and briefly mention lists before moving on the data frames which are the topic of the next module.


## 1. Vectors

```{r}
# These objects hold a sequence of elements of the same data type (e.g., a series of numbers or words/strings)

# We can create vectors using the concatenate function c() to combine different elements together
# For example:
my_favorite_numbers <- c(2, 7, 11, 100, 20)
days <- c("Monday", "Tuesday", "Wednesday")

my_favorite_numbers
days

# We can also create vectors using the repeat rep() function and the sequence seq() function.
# For information and examples, you can run help(rep) and help(seq).
help(seq)
```


## 2. Matrices

```{r}
# Matrices are 2-d data structures that hold elements of the same data type.
# There are indexed by rows and columns. They are created using the matrix function.

mat <- matrix(c(1,2,3,4,5,6,7,8,9), nrow = 3, ncol = 3)
mat

mat_2 <- matrix(c(1,0,0,4,0,0,7,0,0), nrow = 3, ncol = 3)

# Mathematical operations for vectors and matrices occurs elements wise.
mat
mat - mat_2
```


## 3. Data Frames

```{r}
# Data frames are also 2-d data structures with rows and columns.
# However, unlike matrices, data frames can hold elements of different data types.
# They are typically used for data analysis where a column represents a feature of the dataset.

# They are made using the data.frame function and passing in vectors representing 
# the features of our dataset
restaurant_df <- data.frame(weekday = c("Mon", "Tue", "Wed", "Thur", "Fri", "Sat", "Sun"),
                 num_of_waiters =c(2, 2, 2, 3, 5, 5, 3),
                 num_of_cooks = c(2, 2, 2, 2, 4, 4, 2),
                 customers = c(55, 44, 50, 75, 150, 150, 60))
restaurant_df

# Alternatively, we can define the data frame from vectors in our environment.

# Let's make some vectors first.
vec1 <- c("Albert", "Bryan", "Carl", "Dan")
vec2 <- c("Guitar", "Drums", "Piano", "Guitar")
vec3 <- c(4, 3.76, 3.5, 3.9)

Music_students_df <- data.frame(Name = vec1,
                                 Instrument = vec2,
                                 GPA = vec3)
Music_students_df

# As we will see, it is easy to manipulate data frames and compute transformations of our columns.

# This begs the question: If matrices and vectors are data frames with limitations on data types, why use them?
# Answer: They are computationally more efficient. If you are writing algorithms or need linear algebra, you should be working with vectors and matrices. 

# For most data analysis and visualizations, we generally don't need to worry about computing resources and will be using data frames.
```

## Exercise 2

```{r}
# (1) A mathematical operation between a single numeric element and a vector or a matrix will perform that for each element in the vector or matrix.

# i) Try running 1 + mat. 

# ii) Find a numeric variable in your environment. Add that variable to mat and save the 
#     results as an object called new_mat.

# iii) Divide new_mat by a number such that all of its values are between -1 and 1.

#(2)
# i) seq(from = 1, to = 100, by = 1) generates the sequence 1,2, ..., 99, 100.
#    Use seq() to generate the vector -100, -98, -96, ..., 98, 100.
#    Save the results in a vector called plus_minus_100_seq

# ii) Create a vector that holds the numbers 100, 0, 1, 2 ,..., 100.
#     Hint: use the concatenate function c() and the seq() function.
```


## 4. Lists

```{r}
# Lists allows you to store different data types and structures together in a single object

list_example = list(vec = c(1,2,3), string = 'word', matrix = mat)

# We can then access elements of the list with the '$' operator:
list_example$vec
list_example$matrix

# or we can access them by their index number
list_example[[1]] # returns vec
list_example[[2]] # returns string
list_example[[3]] # returns matrix


# "Lists" will come in handy later when we write functions and we want multiple outputs.

```


## Exercise 3

```{r}
# 1) We can check the number of items in a list by passing it to length() function.
# For example running the command length(list_example) returns the number of elements in a given list. 
# Try running length(list_example)


# 2) We can get the names of elements in a list by passing it to names() function.
# Try running names(list_example)
```
